Particle rejuvenation of Rao-Blackwellized Sequential Monte Carlo smoothers for Conditionally Linear and Gaussian models

### 2017-05-15 TexArxiv/answer_referees.tex

The paper surveys recent Rao-Blackwellized particle smoothers and presents a particle rejuvenation approach. The particle rejuvenation is interesting and seems useful, but it also shares some resemblance with the MCMC-based rejuvenation strategy proposed by [LBS+16, Section V.B] in the context of Rao-Blackwellized FFBS and with the strategy proposed by [FWT10] in the context of two-filter smoothing. Please relate the proposed method to these prior works. The detailed comments by the reviewers need to be addressed as well.

\vspace{.2cm}

{\em

It is true that our rejuvenation step shares some resemblance with the algorithms of [FWT10] and [LBS+16, Section V.B]. As noted in [FWT10, Secion~2.6], two-filter smoothers are prone to suffer from degeneracy issues when the algorithm associates forward particles at time $i-1$ with backward particles at time $i$. The authors illustrate this issue in the case where the hidden state is an AR(2) process. To overcome the weakness of such standard two-filter approaches, [FWT10] samples new particles at time $i$ conditional on forward particles up to time $i-1$ and on backward particles from time $i+1$ which are appropriately weighted. This allows to produce new particles at time $i$ and to obtain a SMC approximation whose support is not restricted to the backward particles at time $i+1$ as in the original two-filter algorithm.  The particle rejuvenation proposed in our paper exploits this idea in the specific case of linear and Gaussian models where explicit computations allow to produce an approximation using $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$ with support $\{1,\ldots,J\}$. In addition, it produces a direct approximation of the target distribution based on $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$  without any additional importance sampling steps.

On the other hand, another modification of the FFBS algorithm based on a Markov chain Monte Carlo sampling step was introduced in [LBS+16]. Instead of sampling from  explicitly from (12), [LBS+16] proposed to draw a forward path $a_{1:i-1}$ in $(a^k_{1:i-1})_{1\le k \le N}$ and a sate $a_i$ in $\{1,\ldots,J\}$ according to:
\[
\widetilde{q}(a_{1:i}|a_{i+1:n},y_{1:n}) = \sum_{k=1}^N \widetilde{\vartheta}^k_{i-1}\widetilde{q}(a_{i}|a^k_{1:i-1},a_{i+1:n},y_{1:n})\delta_{a^k_{1:i-1}}(a_{1:i-1})\eqsp,
\]
where $(\widetilde{\vartheta}^k_{i-1})_{1\le k \le N}$ are adjustment multipliers and $\widetilde{q}(a_{i}|a^k_{1:i-1},a_{i+1:n},y_{1:n})$ is a proposal kernel chosen by the user. This means that an ancestral path $a^{\star}_{1:i-1}$ is sampled in $(a^k_{1:i-1})_{1\le k \le N}$ with weights $(\widetilde{\vartheta}^k_{i-1})_{1\le k \le N}$ and $a^{\star}_i$ is sampled from $\widetilde{q}(\cdot|a^{\star}_{1:i-1},a_{i+1:n},y_{1:n})$. Then, the proposed sequence $a^{\star}_{1:i}$ is accepted or rejected using the usual Metropolis-Hastings acceptance ratio. The choice of MCMC rejuvenation has interesting practical consequences as the computation of the acceptance ratio only requires to compute the posterior probability (11) for the proposed sequence $a^{\star}_{1:i}$ while our technique is based on the computation of  (11) for all combinations of sequences $(a^{k}_{1:i-1})_{1\le k \le N}$ and states $a_i\in\{1,\ldots,J\}$. The method proposed in [LBS+16] is computationally less costly as its complexity depends mainly on the number of MCMC samples produced by the algorithm. However it also requires a fair amount of tuning to choose efficient adjustment multipliers and proposal kernels. Our proposed method is computationally more intensive but allows to sample explicitly from (12) without additional tuning parameters. Note that both methods provide similar results for the same computational costs.
These remarks were added at the end of Section~2.2 and Section~2.3.1.
}

@@

\vspace{.5cm}

\noindent Some (more) intuition, about concepts and why certain steps are performed. Especially in the presentation of the RBTF in Section 2.4 I had problems to follow the overall idea. E.g. where do we integrate out the linear states in the backward pass of the original formulation presented by [BDM10]. I would expect this description before the modification of particle rejuvenation is presented. Also, some intuition about the explicit marginalization proposed in (9) would be helpful. Why is it chosen, what is the benefit compared to the original formulation? Currently, one gets easily lost in the derivations without maintaining the overall picture.

\vspace{.2cm}

{\em
\noindent  This comment helped to greatly improved the readability of the paper. As explained in the previous paragraph, Section~2.3.1 is now devoted to the original Rao-Blackwellized two-filter method and Section~2.3.2 to the rejuvenation step. The idea supporting two-filter based algorithms is clearly presented in Section~2.3.1 and most of the computations have also been postponed to the appendix in order to highlight the overall picture and avoid technicalities. Some additional motivations to introduce rejuvenation are given at the end of Sections~2.2 and~2.3.2 (see the answer to the associate editor). Two-filter smoothers are prone to suffer from degeneracy issues when the algorithm associates forward particles at time $i-1$ with backward particles at time $i$. The particle rejuvenation proposed in our paper exploits the idea introduced in [FWT10] in the specific case of linear and Gaussian models where explicit computations allows to produce an approximation using $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$ with support $\{1,\ldots,J\}$ and without any additional sampling steps.
}

@@

\vspace{.5cm}

\noindent Some comparisons in terms of performance, complexity, detailing the superiority (or not) of one method over the other. E.g. what are the pros and cons of RB-FFBS and RBTF. Also some numerical comparisons in the simulation section are expected.

\vspace{.2cm}

{\em
\noindent Even is the paper is not a survey of all Rao-Blackwellized methods for CLGM anymore, some additional numerical experiments were added in Section~3. Each simulation study which was previously given in Section~3 was devoted either to FFBS or two-filter based methods. Now, both algorithms are used and compared to their counterpart with rejuvenation. Our main objective is then to show the improvement of the particle rejuvenation steps for both methods.  This also confirms that FFBS algorithm performs better even with a sharp choice for the  $\gamma_i$'s but also that rejuvenation steps improve both the accuracy and the variance of each method.
}

@@

\vspace{.5cm}

\noindent {\em The minor comments were taken into consideration and typos were corrected. Some specific details are given below.}

@@

\subsubsection*{Referee 2}
\noindent It is well known that SMC algorithms, in their most basic form, are numerically unstable
when used for approximating distributions defined on path spaces of increasing dimension. This is due to the particle path degeneracy phenomenon caused by the resampling mechanism of SMC algorithms. The degeneracy appears to be a potential problem for the estimators discussed in the paper, since the approximation of the
backward kernel (on, e.g., p. 8, line 52) involves quantities such as the mean and the
covariance matrix of the conditional distribution of $z_i$ given the states $a_{1:i}$ and the observations,
evaluated along the particle genealogical paths. I miss a discussion on this. If the effect exists, does the rejuvenation proposed by the authors counteract the same? (I do not think so).

\vspace{.2cm}
{\em
\noindent As explained to the associate editor and to  Referee~1, it is true that particle smoothers are prone to degenerate when they are based on  the genealogy of the forward or backward particles. In addition, in the specific case of two-filter based smoothers, [FWT10, Section~2.6] noted that these algorithms are prone to suffer from degeneracy issues when the algorithm associates forward particles at time $i-1$ with backward particles at time $i$. 
The particle rejuvenation proposed in our paper produces an approximation using $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$ with support $\{1,\ldots,J\}$. In the specific case of linear and Gaussian models, Kalman filtering techniques allow to compute explicitly the approximation of the target distribution at time $i$ based on  $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$.

Therefore, the support of the approximated distribution is not restricted to $(\tilde{a}^k_{i})_{1\le k \le N}$ to overcome particle depletion. Marginalization over the linear states  builds new trajectories of the form $(a^k_{1:i-1},j,\tilde{a}^\ell_{i+1:n})_{1\le k,\ell \le N}$ for all $1\le j \le J$ which have not been sampled by the forward and backward filters. Although this does not overcome the overall degeneracy issue, numerical experiments highlight that this improves the performance of all Rao-Blackwellized SMC smoothers. Additional comments were added at the end of Section~2.2 and Section~2.3.1.
}

@@

\noindent Finally, I miss a discussion on the last sampling operation of the Rao-Blackwellized
two-filter smoother, i.e., the sampling of the mixture obtained when plugging the approximations into (9). This should yield a mixture with $N^2$ components, implying a not innocent computational complexity when N is large.

\vspace{.2cm}
{\em
\noindent The additional computational cost of the rejuvenation step comes from the Kalman filtering steps needed to extend all backward trajectories (built with particles). These steps are much faster than the combination of the forward and backward particles whose complexity grows with $N^2$. However, this combination is also necessary in the original two-filter algorithm of [BDM10] with no rejuvenation. A comment was added in the numerical section.
}

\vspace{.5cm}

\noindent {\em All minor comments were taken into consideration and typos were corrected.}
\end{document}

### 2017-05-15 TexArxiv/clgm.tex
Final modifications for revision

In addition, Figure~\ref{fig:ffbs:var} displays the empirical variance of the estimation for each algorithm to illustrate that both estimators have similar variability. 

>>>>

In addition, Figure~\ref{fig:ffbs:var} displays the empirical variance of the estimation for each algorithm.
Figure~\ref{fig:ffbs:err} and  Figure~\ref{fig:ffbs:var} illustrate that in both cases the additional rejuvenation step improves the accuracy and the variability of SMC smoothers. In addition, even with a sharp choice for the artificial distributions $\gamma_i$, $1\le i\le n$,  FFBS based methods outperform two-filter based smoothers. 
